---
title: 'Disclosing funding sources for open access publication fees: the Open APC initiative'
author:
- Najko Jahn, Bielefeld University Library, Bielefeld University
- <a href="http://orcid.org/0000-0002-5111-2788">Marco Tullney</a>, Technische Informationsbibliothek (TIB) - German National Library of Science and Technology, Hannover, Germany
date: '7 April 2016'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: peerj.sty
  html_document:
    fig_caption: yes
    force_captions: yes
    number_sections: no
    theme: readable
bibliography: literatur.bib
csl: frontiers.csl
abstract: Publication fees in open access publishing hold a prominent place on the agenda of researchers, policy-makers, and academic publishers. This paper contributes to the evolving empirical basis on open access funding. It describes the Open APC initiative, in which German universities and research organizations share their expenditures for publication fees. As method, the initiative uses existing open data tools to aggregate and disseminate institutional spending on open access publication fees. In total, 29 German research organizations self-reported funding of 6,279 open access journal articles, which amounted to 8,039,339 €. The average payment for each article was 1,280 €, and the median payment 1,209 €. Our data-set comprises only 53 articles in hybrid journals. With an indexing coverage of 99 %, the findings reveal that the DOI agency CrossRef provides both comprehensive bibliographic coverage of the funded open access journal literature and disambiguated names of journal titles and publishing houses. We show that authority control of these bibliographic information is particularly relevant for the comparative study of the economical effects of open access publishing.
---

```{r, echo =FALSE}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 12,
  fig.height = 8
)
options(scipen = 999, digits = 2)
knitr::knit_hooks$set(inline = function(x) {
      if(is.numeric(x)){
          return(prettyNum(x, big.mark=","))
      }else{
          return(x)
       }
   })
```

# Introduction{.unnumbered}

Publication fees, often paid by funders or universities, are a widely discussed open access business model. Yet, how and to what extent these activities are effective in terms of the number of supported research articles and associated costs remains under debate. This paper describes the Open APC initiative[^2], in which German universities and research organization share spendings on open access publication fees, and how it is currently implemented. More specifically, it addresses three problem areas when studying the economical effects of open access publishing: fragmentation of open access funding, variable pricing schemes and the comparison across research institutions. Such an approach extends methods and improves data collection activities for researchers and practitioners, as well as contribute to a better understanding of factors affecting the analysis of publication fees in open access publishing.

The rise of open access journals  matches the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover these fees, authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. However, collecting information about what was funded is in most cases difficult. One reason why payments made for open access journal publications are often hard to track is that, on the one hand, they are fragmented across the budgets of funding agencies, research institutions, and libraries, and, on the other hand, taken from personal budgets. Furthermore, open access funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets stand in contrast with those support structures and are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015]. Along with the fragmentation of payments, funding for open access publications lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014], a situation similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. It also remains unclear which factors contribute to price formation.[^4b] While fixed prices for individual articles are common, agreements between publishers and institutions often provide discounts and publishers sometimes waive publication fees for authors from low-income countries [@BJ_RK_2012; @Lawson_waiver_2015]. Other factors leading to a complex landscape of variable pricing schemes [@Pinfield_2015] include submission or page charges [@BJ_RK_2012]. Hybrid journals substantially add to this complexity, because comprehensive offset systems to avoid paying for the same article twice, through subscription and publication fee, are rare, which, in turn, leads to the phenomena of "double dipping" in scholarly publishing [@Pinfield_2015].

This complex situation of fee-based open access publishing creates difficulties for researchers and practitioners alike. Because of fragmented payments, the extent of funding remains unclear. To increase transparency, some research funders have begun collecting and disclosing expenditures for open access journal articles as open data. As per definition, open data is data that "can be freely used, modified, and shared by anyone for any purpose" [@OpenDataHandbook]. Therefore, opening up information about the funding of open access journal articles promises to enhance the discussion about current and future business models in academic publishing. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc] and the Austrian Science Fund FWF [@fwf_apc], who both released data on publication fees they had funded. The British not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report expenditures along with bibliographic information, including title, journal and publisher, persistent identifier to the publisher's version, and a link to a deposit in a subject repository. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates. In the case of the Wellcome Trust, crowd-sourcing data cleaning activities through a Google spreadsheet in combination with checks against bibliographic sources massively improved the spending data (see comments in @wellcome_apc).

The open access landscape in Germany, which is the focus of this paper, shares the general problems of in-transparency regarding funding schemes and costs as discussed above. The Deutsche Forschungsgemeinschaft (DFG), the largest research funder in Germany, has been encouraging open access publishing since years. It launched its "Open-Access Publishing" program in 2009 that has strongly influenced the support of open access publication fees through funds managed by university libraries.[^3] With this program, the DFG aims to help universities to establish support structures for publishing in open access journals where authors are requested to pay a publication fee. To reduce administrative burdens, grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. Examples include central invoicing schemes and related agreements between university libraries and publishers. Grantees are also required to report the institutional publication output and their fees paid for open access journal articles to the DFG on a regular basis, and to present the university-wide strategy to sustain the funds when DFG’s initial support runs out lately in 2019. The DFG enforces a set of criteria grantees have to comply with, leading to similar implementations for supporting open access publishing across German universities: these criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding VAT) [@Fournier_2013]. Research institutes organized in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft are not eligible for this funding program, contributing to the diversity of schemes in Germany. In response, some organizations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The growing share of articles published in fee-based open access journals in recent years has led to calls for an unified approach towards funding of publication fees. The Allianz der Wissenschaftsorganisationen[^10c], representing all major research organizations in Germany, thus marks transparency as a major means to sustain an "adequate open access publication system" [@allianz]. However, there are various ways to achieve this goal. The existing approaches in Austria and the United Kingdom have one institution in charge to collect and analyze the data. The history of the Open APC initiative is rather bottom-up: In May 2014, Bielefeld University Library began to share its expenditures for publication fees. The library put its approach to the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11] as a basis for discussion, and invited others to participate. Reflecting the increasing demand for publicly available data, contributions from Universität Regensburg and Universität Hannover followed soon after. As of writing, 29 universities and research institutes voluntary reported their data to the Open APC initiative to be included into a unified data-set of all expenditures.

Drawing on self-reported cost data, we asked to what extend German universities and research institutions covered publication fees for open access journal articles over the last years. More specifically, we addressed three questions that arose while studying fee-based open access publishing:

- how to gather bibliographic metadata about sponsored open access journal articles,
- how is institutional publication fee spending distributed over publishers and journal titles,
- what is the share of sponsored articles in open access journals versus hybrid journals.

# Methods and Materials{.unnumbered}

The study examined cost information provided by the Open APC initiative. Under this umbrella, German universities and research institutions shared details of how much open access journals charged authors as open data. We analysed the data-set that was released on 13 May 2016 and is stored as version v2.4.2 in the corresponding GitHub repository. At the time of our study, the data-set reported the amount institutions paid for each article including value added tax, and the year an article was invoiced. Bibliographic metadata consisted of the Digital Object Identifier (DOI) to uniquely identify an article, the publisher, and information about the journal titles. The Open APC initiative automatically retrieved this publisher and journal information from CrossRef, a DOI minting agency for scholarly literature. Furthermore, the data-set differentiated whether an open access article was published in a subscription based journal ("hybrid journal").

For gathering normalized publisher and journal titles, we re-aggregated the bibliographic metadata from CrossRef using the DOI on 19 May 2016 because of two reasons: First, CrossRef provides incremental metadata updates to reflect the dynamic landscape of academic publishing including ongoing mergers of publishing houses or name changes. A prominent example in this regard is the merger of the two large publishing houses Springer Business + Media and Nature Publishing that now operate under "Springer Nature". In contrast to this, the Open APC initiative normalized publisher and journal titles based on CrossRef shortly after the institution had provided its cost data. Therefore, re-aggregating the data-set ensures unique reference to currently existing publishers and journals at the expense of information about to which publisher a journal belonged while a sponsored article was published.

Secondly, CrossRef is not the only registration agency for DOI's, but also DataCite and Medra mint DOIs for scholarly work. While obtaining bibliographic information from CrossRef, we also examined the number of publications for which each of the DOI registration agency provided metadata during our study. As a client, we used the R package rcrossref [@rcrossref] developed and maintained by the rOpenSci initiative.[^6] With the function `cr_cn`, we requested the linking resource `application/vnd.crossref.unixsd+xml`, which main function is to support text mining activities.[^5] The advantage of this XML format was that it distinguishes full and abbreviated journal titles as well as the media types of ISSN's, the International Standard Serial Number used to identify journals. It also contained license information and disambiguated publisher information, thus avoiding confusion about licensing and naming of publisher houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values.

After summarizing publication fee spending per institution, publisher and journal, we compared the share of hybrid journal articles in the Open APC data-set with those of Wellcome Trust, FWF and Jisc. For this aim, we systematically reviewed the proportions for each year from the accompanying literature. 

# Results{.unnumbered}

```{r, echo=FALSE, cache = FALSE}
# load data
my_apc <- read.csv("data/apc_de.csv", header = TRUE, sep =",", stringsAsFactors = FALSE)

# data cleaning
library(dplyr)
# remove Bielefelds Zero value
my_apc <- filter(my_apc, euro > 0)
# remove intech book chapters
my_apc <-  filter(my_apc, !is.na(journal_full_title))
# clean publisher titles, because metadata could not be fetched
my_apc$publisher[grep("BioMed Central", my_apc$publisher, ignore.case = T)] <- "Springer Science + Business Media"
my_apc$publisher[grep("Impact Journals, LLC", my_apc$publisher, ignore.case = T)] <- "Impact Journals LLC"
my_apc$publisher[grep("e-Century Publishing Corporation", my_apc$publisher, ignore.case = T)] <- "e-Century Publ. Corp."
my_apc$publisher[grep("Academic Journals", my_apc$publisher, ignore.case = T)] <- "Academic Journals"
# fix plos name change in journal titles
my_apc$journal_full_title <- gsub("PLoS", "PLOS", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("PloS","PLOS", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("PlOS","PLOS", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("Sustainibility","Sustainability", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("Surgeal","Surgical", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("Oncotarget","OncoTarget", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("BMC cancer","BMC Cancer", my_apc$journal_full_title)

```

```{r results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(scales)

```

## Cost Data{.unnumbered}

```{r}
apc_time <- my_apc %>%
  group_by(period) %>%
  summarize(articles = n())
```

On February 19th 2016,[^16a] the Open APC initiative covered `r nrow(my_apc)` articles, whose publication fees were centrally paid by `r length(unique(my_apc$institution))` German universities and research institutions. The number of supported open access journal articles, which were reported to this initiative, grew over the years (see Figure 1). While one institution disclosed `r filter(my_apc, period == min(my_apc$period)) %>% nrow()` payments made in `r as.character(min(my_apc$period))`, the majority shared their expenditures from 2013 onwards. With `r max(apc_time$articles)` articles, the year `r apc_time[apc_time$articles == max(apc_time$articles),"period"]` was best represented in our data-set. Because of a time lag between payments made and reporting them to the Open APC initiative, only `r my_apc %>% filter(period == 2015) %>% group_by(institution) %>% summarise(n()) %>% nrow()` institution were able to partly contribute their cost data for 2015 at the time of this analysis.

```{r articles_year, fig.width=6, fig.height=2.5,fig.cap="Growth of Open APC Initiative"}
apc_time <- my_apc %>%
  group_by(period) %>%
  summarize(articles = n())
ggplot(apc_time, aes(factor(period), cumsum(articles), group = 1)) +
  geom_line(stat = "identity", colour = "blue") +
  geom_point() +
  xlab("Year") +
  ylab("OA articles reported\n(cumulative sum)") +
  scale_y_continuous(limits = c(0,8000))
```

```{r}
iqr_df <-  my_apc %>%
  group_by(period) %>%
  summarise(iqr = mean(euro))
```

Among all articles, fees amounted to `r sum(my_apc$euro)` € including VAT, the average payment was `r mean(my_apc$euro)` € and the median value `r median(my_apc$euro)` € . Figure 2 shows the large price variation among the articles. The disclosed publication fees ranged from `r min(my_apc$euro)` € to `r max(my_apc$euro)` €. However, the average price paid varied somewhat during the period 2011 and 2014 (`r my_apc %>% filter(period == 2011) %>% summarise(mean(euro)) %>% round()` - `r my_apc %>% filter(period == 2014) %>% summarise(mean(euro)) %>% round()` €). We also observe that `r nrow(my_apc[my_apc$euro < 2000,])` (`r (nrow(my_apc[my_apc$euro < 2000,]) / nrow(my_apc))  *100`%) of the publication fees were paid in accordance with the DFG price cap of 2,000 €. Whereas related open data initiatives in Austria and the United Kingdom reported a large share of spending for hybrid journal articles, the situation in Germany is different: only  `r my_apc %>% filter(is_hybrid == TRUE) %>% nrow()` articles in hybrid journals were reported by `r my_apc %>% group_by(institution) %>% filter(is_hybrid == TRUE) %>% summarise() %>% nrow()` out of `r length(unique(my_apc$institution))` research institutions, accounting for `r (my_apc %>% filter(is_hybrid == TRUE) %>% nrow() / my_apc %>% nrow()) * 100` % of the overall payments.


```{r payments_year, fig.width=6, fig.height=3,fig.cap="Payments per year"}
ggplot(my_apc, aes(factor(period), euro)) +
  geom_boxplot() +
  xlab("Year") +
  ylab("Publishing fee in EURO(log scale)") +
  scale_y_continuous(limits = c(0,8000))
```

The number of APC payments per institutions varied considerably (see Table 3). With `r filter(my_apc, institution == "MPG") %>% nrow()` reported articles, the Max Planck Society contributed `r (filter(my_apc, institution == "MPG") %>% nrow() / nrow(my_apc)) * 100` % of the overall submissions. In contrast, the two universities of technology, TU Clausthal and TU Ilmenau, who recently begun to set up support structures for fee-based open access journal articles, shared payments made for four articles each.


```{r results='asis'}
tt <- my_apc %>%
  group_by(institution) %>%
  summarise(Articles = n(), Total = sum(euro), Mean = mean(euro), Median = median(euro), Min = min(euro), Max = max(euro)) %>%
  arrange(desc(Articles))

colnames(tt) <- c("Institutions", "Articles funded", "Total", "Mean", "Median", "Minimum", "Maximum")
tt <- format(tt,big.mark=',')
pander::pandoc.table(tt, split.cells = 7, caption = 'Institutions self-reporting expenditures for open access publications (in €)', justify = 'lrrrrrr')
```

## CrossRef indexing{.unnumbered}

Along with the price information, participating institutions were required to identify funded articles by their DOI. They were reported for `r my_apc %>% filter(!is.na(doi)) %>% nrow()` out of `r nrow(my_apc)` articles. Of those, `r my_apc %>% filter(indexed_in_crossref == TRUE) %>% nrow()` were indexed in CrossRef, representing `r my_apc %>% filter(indexed_in_crossref == TRUE) %>% nrow() / my_apc %>% nrow() * 100`  % of all funded publications. The reasons why articles identified by a DOI were not registered with CrossRef differed. Some journals were not indexed by CrossRef at the time of our study but by the DOI agencies DataCite (Journal of new frontiers in spatial concepts published by KIT Scientific Publishing) and Medra (DIE ERDE: Journal of the Geographical Society of Berlin). In other cases, either the DOI did not refer to the full text despite the fact that the journal was indexed on a regular basis (compare <http://doi.org/10.1186/1471-2105-13-S19-S7> with <http://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-13-S19-S7>) or the resource type could not be retrieved, although the DOI resolves (<http://doi.org/10.1186/s12885-015-1795-7>).

## Cost data by publisher and journal{.unnumbered}


```{r}
apc_publisher <- my_apc
apc_publisher$publisher <- factor (apc_publisher$publisher, levels = c(rownames(data.frame(rev(sort(table(apc_publisher$publisher)))))))

levels(apc_publisher$publisher)[11:length(levels(apc_publisher$publisher))] <- "other"
```

We used the DOI to automatically fetch publisher and journal names for each article from the CrossRef REST API. Table 4 shows the top ten publishers in terms of payments made that represent `r apc_publisher %>% filter(!publisher == "other") %>% nrow() / my_apc %>% nrow() * 100` % of the spending for publication fees. In total, payments were made to  `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. In comparison with data from the UK, full open access publishers have a greater share on total spending. @Pinfield_2015, for instance, reported remarkably lower numbers for the open access publishers MPDI AG, Copernicus GmbH, and Hindawi Publishing.

```{r results='asis'}
tt <- apc_publisher %>%
  group_by(publisher) %>%
  summarise(Articles = n(), Total = sum(euro), Mean = mean(euro), Median = median(euro), Min = min(euro), Max = max(euro))  

colnames(tt) <- c("Publisher", "Articles funded", "Total", "Mean", "Median", "Minimum", "Maximum")
tt <- format(tt,big.mark=',')
pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per publisher (in €)', justify = 'lrrrrrr')
```

Most of the funding of publication fees in Germany went to the publisher Springer Science + Business Media, especially profiting from the merge with the former full open access publisher BioMed Central. In contrast, other established publishing houses such as Elsevier and Wiley-Blackwell rank lower, presumably because they mostly publish hybrid journals, which were not well represented in our data-set at the time of the study. Table 4 also illustrates the variation across and within publishers, which confirms earlier findings [@Pinfield_2015].


```{r results='asis'}
apc_journal_full_title <- my_apc
apc_journal_full_title$journal_full_title <- factor (apc_journal_full_title$journal_full_title, levels = c(rownames(data.frame(rev(sort(table(apc_journal_full_title$journal_full_title)))))))

levels(apc_journal_full_title$journal_full_title)[11:length(levels(apc_journal_full_title$journal_full_title))] <- "other"

tt <- apc_journal_full_title %>%
  group_by(journal_full_title) %>%
  summarise(Articles = n(), Total = sum(euro), Mean = mean(euro), Median = median(euro), Min = min(euro), Max = max(euro))

colnames(tt) <- c("Journal", "Articles funded", "Total", "Mean", "Median", "Minimum", "Maximum")
tt <- format(tt,big.mark=',')
pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per journal (in €)', justify = 'lrrrrrr')
```

Prices also varied within single journals. Based on the number of articles paid for, Table 5 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals. Payments to these ten journals represent `r apc_journal_full_title %>% filter(!journal_full_title == "other") %>% nrow() / my_apc %>% nrow() * 100` % of all payments. In the case of Atmospheric Chemistry and Physics Discussions, the price range can be explained by the fact that this journal charges per page and also takes the submission's file format into consideration.

The data-set finally confirms the leading role of "mega-journals" in open access publishing, including the multidisciplinary PLOS ONE and the journals New Journal of Physics, Atmospheric Chemistry and Physics Discussions and Frontiers in Psychology, all of which publish contributions from all branches of their respective discipline. In general, an estimated 14 out of more than 10,000 journals registered in DOAJ in 2015 accounted for up to 15–20 % of all articles published in full open access journals [@Bj_rk_2015].

# Discussion{.unnumbered}

The Open APC initiative extends existing methods to disclose spending on open access publication fees. Our workflow benefits from openly available tools and the social coding platforms GitHub, both of which are well established and suited to increase transparency in research [@Peng_2011; @Ram_2013]. For 99 % of the articles, CrossRef provided bibliographic information, which substantially contributed to a uniform data-set about formerly fragmented payments made for open access articles.

Although CrossRef disambiguates journal titles and publisher names and is therefore an authority-controlled source for open access journal literature, derivations from CrossRef metadata curation as well as the context of aggregation must be made clear. In particular, problems persist on how to deal with name changes and ongoing mergers. For example, the publisher Public Library of Science (PLOS) has changed its acronym from "PLoS" to "PLOS", which CrossRef metadata reflects from 2015 onward. We therefore normalized all PLOS journal titles in order to secure unique reference to these journals. Another publisher affected is "The Optical Society," formerly "Optical Society of America". Because the ownership of publishing houses can become combined, dealing with mergers is also essential to make cost data comparable. Jisc data, for instance, differentiate between the full open access publisher BioMed Central and the traditional publisher Springer, concluding that "traditional publishing houses" lean on the hybrid model [@Pinfield_2015]. This stands in stark contrast to our approach, in which CrossRef metadata reflects the merger of BioMed Central and Springer, resulting in Springer Science + Business Media to be the best represented publisher for articles in full open access journals in the Open APC data-set. Another approach for ensuring unique reference, but that we have not evaluated yet, is to use CrossRef's identifiers for journals and publishers instead.[^19]

Because of the dynamic landscape of academic publishing and its representation in CrossRef's data curation efforts, it is important to consider the time-frame of metadata aggregation. In our case, we re-used metadata shortly after the data submission. However, for some cost analysis -- for instance to prepare negotiations with publishers on future schemes to fund open access journal articles -- it could be more feasible to re-normalize the complete data. The rossRef API provides incremental metadata updates that can be used to assess the current potential of future funding. While licensing information is incompletely covered in the CrossRef index so far, and therefore not analyzed in our study, the growing importance of facilitating text mining may result in more and more publishers sharing this information with CrossRef in the future.

Participation is voluntary. Therefore, not all institutions in Germany that provide central funding of publication fees contribute cost data to this initiative. In a qualitative survey that also asked why German institutions are reluctant to share their cost data through the Open APC initiative one institution feared that increase in transparency would allow publishers to adjust prices in their favor. Others pointed out that the workload to produce such a data-set could be too extensive [@deppe_2015].

While there may still be institutions that have no overview of their APC spending, we would like to emphasize that reporting data that is already available within an institution to the Open APC initiative should not lead to much additional work. The central incentive of this initiative is to make it as easy as possible to submit data to it. Being able to combine that data into one standardized data-set increases transparency and comparability, and gives institutions a better understanding of the overall development of open access publishing. In particular, we cannot see the harm of increased transparency; in fact, not knowing how much is spent is undoubtedly a disadvantage in dealing with publishers.

Extending the data template to include information about funders or whether special agreements with publishers applied as suggested by @Pinfield_2015, would even increase the efforts needed to participate. However, with the growing demand for action in areas like the large-scale transition of toll-access journals to open access [@Schimmer_Geschuhn_Vogler], an updated data template could help institutions to better comply with these policy developments in the future. From our experience, another barrier to participate is the lack of skills in version control: the data submission itself is not always made directly by the institutions. Instead, they sent files to Bielefeld University Library with the request to make them available on GitHub on their behalf.

Future work needs to focus on analyzing the cost data. Of particular interest are questions concerning the coverage of central funding schemes in comparison with open access publication output in general, and the use of other means to cover publication fees in particular. The Open APC initiative does not cover personal budgets or make price reductions explicit, but studies suggest that there is a possible gray area [@BJ_RK_2012; @Bj_rk_2015; @Lawson_waiver_2015]. Existing study designs could be re-applied to examine the relationship between price on the one hand, and indexing coverage, journal prestige or management costs on the other [@Walters_2011; @Bj_rk_scientometrics_2015, @Pinfield_2015]. This, in turn, helps to address the central question of future business models in scholarly publishing from an international perspective [@Pinfield_2015].

<!-- # Conclusion{.unnumbered}

The purpose of the Open APC initiative is to gain better insight on spending for open access. The main operational idea—to make data submission easy by building upon existing data and by using popular open tools and data bases—has proven to be a feasible approach and to scale well. Starting with a few institutions, the integration of further data contributors has not been a problem. While we use article metadata reported by publishers to CrossRef, we think it's best to collect information on how much has been spent at the sources: the institutions paying for APCs. We look forward to more data contributors to the initiative and to adapt this scheme to new analyses, additional data, or international sources. The Open APC initiative can serve as an example of the value that open data, open source software, and the knowledge and skills of tech-savvy librarians add to modern tasks of libraries.-->



# Acknowledgment{.unnumbered}

We thank Andrea Hacker and Ada-Charlotte Regelmann for valuable comments on the first draft of this paper. We also thank Christoph Broschinski, Vitali Peil, and Dirk Pieper, the members of the DINI working group "Electronic publishing", and all data contributors[^20] of the Open APC initiative.


# References{.unnumbered}


[^1]: Comprehensive list of business and revenue models of open access journals: <http://oad.simmons.edu/oadwiki/OA_journal_business_models>

[^2]: <https://github.com/openapc/openapc-de>

[^2a]: Project summary available at <https://github.com/OpenAPC/openapc-de>. The initiative is now supported by the INTACT project: http://www.intact-project.org/.

[^2b]: However, it should be noted that the administrative costs of processing invoices for a large number of articles are not measured in this project.

[^3]: Guidelines for the funding program can be found here: <http://www.dfg.de/formulare/12_20/>

[^3a]: We are sure that some universities with DFG-funded publication funds pay for APCs higher than 2,000 EUR. They have to make sure they don't use the funder's money for these articles, and this maybe explains why some institutions are reluctant to report on these APCs. Scatter plots of all APC payments show that there are only very few APCs over 2,000 EUR.

[^4b]: These might include article processing, impact, rejection rates, management and investment, and profit margins. See @Van_Noorden_2013 for a general discussion and @Gumpenberger_2012 and @Bj_rk_scientometrics_2015 for discussions of journal impact and quality.

[^5]: CrossRed: Text and Data Mining for Researchers: <http://tdmsupport.crossref.org/researchers/>

[^6]: rOpenSci: <https://ropensci.org/>

[^7]: Europe PubMed Central RESTful Web Service: <https://europepmc.org/RestfulWebService>

[^8]: Thomson Reuters Article Match Retrieval Service: <http://wokinfo.com/directlinks/amrfaq/>

[^9]: <https://doaj.org/faq#metadata>

[^10]: <http://rmarkdown.rstudio.com/>

[^10c]: <http://www.dfg.de/en/dfg_profile/alliance/index.html>

[^11]: <http://dini.de/english/ag0/e-pub0/>

[^12]: <http://openapc.github.io>

[^13]: <https://jekyllrb.com/>

[^13a]: <http://opendatacommons.org/licenses/odbl/1-0/>

[^14]: <https://lists.uni-bielefeld.de/mailman2/cgi/unibi/listinfo/open-apc>

[^15]: <https://github.com/OpenAPC/openapc-de/wiki>

[^16]: <https://github.com/OpenAPC/openapc-de/issues>

[^16a]: The data is openly available on GitHub. The following analysis is based on version 2.1.13 of the dataset, available at <https://github.com/OpenAPC/openapc-de/tree/v2.1.13>.

[^17]: <http://contributor-covenant.org/>

[^18]: <https://github.com/CrossRef/rest-api-doc/blob/master/rest_api.md>

[^19]: We would like to thank Martin Fenner for pointing this out to us. See also the CrossRef API documentation <https://github.com/CrossRef/rest-api-doc/blob/master/rest_api.md>

[^20]: <https://github.com/OpenAPC/openapc-de#contributors>

[^22]: <http://dx.doi.org/10.4119/UNIBI/UB.2014.18>

[^23]: <https://github.com/OpenAPC/openapc-de/wiki/Handreichung-Dateneingabe>
